% ---------------------------------------------------------------------------- %
% INTRODUCTION
% ---------------------------------------------------------------------------- %
\section{Introduction} \label{sec:intro}
An Ordered Boolean Decision Diagram (OBDD) is a directed acyclic graph that
represents a boolean function in a compact and canonical form. Each node
represents a boolean variable $x_i$ and has two outgoing arcs corresponding to
the two values of $x_i$. The evaluation, composition and manipulation of boolean
functions in an OBDD translates into graph traversal and manipulation
algorithms. By exhaustively applying reduction-rules, one can ensure the graph
is the minimal representation of a boolean function \cite{Bryant86}. The use of
OBDDs have become a core stable in symbolic model checkers such as \emph{MCK}
\cite{Gammie04} and \emph{MCMAS} \cite{Lomuscio17}, circuit design \todocite,
and other applications of Software and Hardware Verification.

A lot has been done to minimise the memory footprint and optimise the speed of
this data structure, such as the work of \textcite{Karplus88} and
\textcite{Brace90} back in the late 80s and early 90s. Still to this day,
attempts are made to further optimise OBDDs, such as the work of
\textcite{Dijk16}. A common idea deployed in these solutions is to add a
memoisation table to the recursive algorithm. This not only stops unneeed
recomputation, but at the same time allows to ensure the minimal size of the
OBDD by directly applying the reduction rules as part of the lookup. In
practice, the memory hierarchy creates a bottleneck on this memoisation table,
since every lookup in the table will result in a cache miss when the data
structure becomes too large. Furthermore, the use of such of such a table across
multiple manipulations results in needing to do garbage collection to remove dead
nodes \cite{Karplus88,Brace90,Dijk16}.

In 1996, \textcite{Arge96} proposed to use a different approach to compute the
manipulations of OBDDs. Using time-forward processing one can rewrite a
recursive algorithm into a non-recursive iterative one, where the recursion is
postponed until the needed data is in memory. A description was given of the two
basic OBDD algorithms \Reduce\ and \Apply\ that only uses the I/O sorting bound
in the worst case.

We follow up on Arge's work, by extending the idea of using time-forward
processing to the other commonly used algorithms for OBDDs and identifying
multiple ways to optimise the constants in them.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% ispell-local-dictionary: "british"
%%% End:
